Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-22 paper.qmd
Update paper.qmd

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  Reproducible analysis requires more than clean, well-documented code: 
  researchers must also manage software dependencies, computational environments, and workflow execution. 
  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles multi-language environments, 
  system dependencies, and pipeline orchestration together. I present a two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each 
  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking. 
  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports 
  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.

@@

## Introduction: Reproducibility is also about software {#secntro}

@@

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, we interpret this notion more broadly as
the *computational environment*: the complete set of software required to
execute an analysis. Here too, a continuum exists. At the minimal end, authors
might only name the main software used—say, the [R]{.proglang} programming
language. More careful authors might also specify the version of [R]{.proglang},
or list the additional packages and their versions. Rarely, however, do authors
specify the operating system on which the analysis was performed, even though
differences in operating systems can lead to divergent results when using the
same code and software versions, as shown by @neupane2019. It is even less
common for authors to provide step-by-step installation instructions for the
required software stack.

@@

This paper focuses specifically on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. I present a comprehensive framework that addresses both
challenges using the [Nix]{.pkg} package manager, making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing our approach, I survey existing tools and
their limitations to contextualize our contribution.

@@

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[[renv]{.pkg} documentation](https://rstudio.github.io/renv/articles/renv.html#caveats).

@@

These tools represent significant progress in reproducibility, but they share a
common limitation: **none addresses system-level dependencies or supports
multi-language environments**. A project using [R]{.proglang},
[Python]{.proglang}, and system tools like [Quarto]{.pkg} or \LaTeX requires
coordinating multiple package managers ([renv]{.pkg} for R, virtual environments
for Python, manual installation for system tools), each with its own
configuration and potential for conflicts.

@@

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. Containers are typically run
nonnteractively, allowing the complete computational environment to be
instantiated and executed with a single command.

@@

1. Poor interactive support: While containers can be modified at runtime,
changes are lost when the container stops unless explicitly saved. Running
graphical applications like RStudio Desktop from containers requires complex X11
configuration that works reliably only on Linux. Instead, web-based IDEs (like RStudio
Server) are used instead, but come with the inherent limitations of web-based solutions.

@@

4. Single-language focus: While [Docker]{.pkg} can support multi-language
environments, orchestrating them requires manual coordination. There is no
standard way to specify that one step should run in a Python environment while
another runs in R.

@@

However, achieving this required mastering multiple complex tools: [renv]{.pkg}
for package management, [Docker]{.pkg} for environment containerization, and
[Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
and failure modes. The complexity compounds for multi-language projects—a Python
project would require learning virtual environments or conda, and coordinating R
and Python in a single analysis requires additional tooling.

@@

Together, these packages provide what existing tools cannot: a single framework
for managing reproducible, multi-language computational environments and
executing complex analytical workflows with step-level isolation. While
[Nix]{.pkg} remains complex for advanced use cases, [rix]{.pkg} and
[rixpress]{.pkg} abstract this complexity for common research workflows, making
deep reproducibility accessible without requiring systems administration
expertise.

The remainder of this paper proceeds as follows. Section 2 introduces the
[Nix]{.pkg} package manager and explains why its functional approach enables
reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
definition for various use cases. Section 4 discusses our `rstats-on-nix` fork
of the official package repository and how it addresses practical limitations.
Section 5 introduces [rixpress]{.pkg} and demonstrates polyglot pipeline
orchestration through a complete example. Section 6 concludes with discussion of
appropriate use cases and future directions.

@@

[Nix]{.pkg} is a powerful, cross-platform package manager designed for installing 
and building software in a fully reproducible and reliable manner. Unlike traditional 
package managers, [Nix]{.pkg} emphasizes immutable infrastructure and a functional
approach, which significantly enhances the consistency of computational environments.
As of writing, the primary Nix package collection, `nixpkgs`, provides access to over 
120,000 packages, including nearly all of CRAN and Bioconductor, making it the largest,
and among the most up-to-date package repository according to Repology:

![nixpkgs is the largest and among the most up-to-date packages repository](repology.png){#fig-repology}

@@

For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, we refer readers to @dolstra2004nix.

By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of 
tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files 
and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
multi-language environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
This enables the creation of a truly complete, project-specific, and deeply reproducible environment
that can be used interactively for development or nonnteractively for automated analysis. 
As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
the environment can be rebuilt reliably in the future.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

Writing [Nix]{.pkg} expressions can be challenging for users unfamiliar with the
[Nix]{.proglang} language. However, the ability to define a fully reproducible
development environment in a single text file and then rebuild it anywhere is
highly appealing. To lower the barrier to adoption of [Nix]{.pkg} for
reproducibility, we developed the [rix]{.pkg} package.

@@

As explained earlier, [Nix]{.pkg} uses expressions from the `nixpkgs` GitHub
repository to build software. However, when generating expressions with
[rix]{.pkg}, our fork `rstats-on-nix/nixpkgs` is used instead.

@@

I strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. We
backported the fix to our fork, making RStudio available to users of our fork
earlier than upstream, as merging PRs in the official repository can take some
time.

@@

A drawback of forking `nixpkgs` is that backported packages are not included
upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
to build many packages from source, which can be time-consuming. To mitigate
this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
complementing the public Nix cache. Instructions for using Cachix are in
[rix]{.pkg}’s documentation. Using the cache significantly speeds up
installations, as prebuilt packages are downloaded rather than compiled.

@@

To illustrate these capabilities with a realistic research task, we present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code to this example can be found on 
[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
using functions inspired by [targets]{.pkg}. The underlying logic for each step
is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
`functions.R`). Further details are detailed step-by-step instructions are provided
in the Appendix. The high-level orchestration script demonstrates how `{rixpress}` defines a
granular, multi-step machine learning workflow that crosses language boundaries:

@@

As a sidenote: [Python]{.proglang} users that wish to use [ryxpress]{.pkg} 
would define the pipeline exactly the same way, as an [R]{.proglang} script.
[ryxpress]{.pkg} calls [R]{.proglang} and [rixpress]{.pkg} under the
hood to generate the `pipeline.nix` file and then build it.

@@

To execute the pipeline, one can either set `build = TRUE` in `rxp_populate()`
or call `rxp_make()` separately. [Nix]{.pkg} executes each step in order,
building dependencies as needed. Outputs are cached, so subsequent runs only
recompute steps with changed inputs or code. This provides the efficiency of
[targets]{.pkg} with multi-language support and bit-for-bit reproducibility.
Artifacts can be inspected interactively in [R]{.proglang} using
`rxp_read("artifact_name")` or `rxp_load("artifact_name")`.

@@

However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
by providing a familiar interface and workflow. By building reproducible
development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
applications, or serving [plumber]{.pkg} APIs.

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of multi-language analyses, making it easier to manage complex
projects with confidence that results can be reproduced exactly.

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide significant steps forward, it is 
important to acknowledge current limitations. Both packages are relatively recent. 
[rixpress]{.pkg} (and [ryxpress]{.pkg}), in particular, requires further development, especially to enhance
the ease of debugging complex pipelines. Its visualization capabilities are also still
fairly limited; for instance, they do not currently indicate outdated derivations when
a user modifies code that would necessitate a rebuild. These areas represent important 
avenues for future work to further mature the framework and improve the user experience.

@@

*   `functions/`: This directory contains the helper scripts with the core analytical code for each language.
    *   `functions.jl`: The [Julia]{.proglang} script for the economic simulation.
    *   `functions.py`: The [Python]{.proglang} script for the machine learning workflow.
    *   `functions.R`: The [R]{.proglang} script for data visualization.
*   `gen-env.R`: An [R]{.proglang} script that defines the reproducible computational environment.
*   `gen-pipeline.R`: The main [R]{.proglang} script that defines and orchestrates the entire multi-language pipeline.

>>>>

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  Reproducible analysis requires more than clean, well-documented code: 
  researchers must also manage software dependencies, computational environments, and workflow execution. 
  Existing tools like [Docker]{.pkg} and [renv]{.pkg} address parts of this challenge, but no single solution handles polyglot environments, 
  system dependencies, and pipeline orchestration together. I present two [R]{.proglang} packages that leverage [Nix]{.pkg} to provide both reproducible polyglot 
  environments and declarative workflow orchestration. [rix]{.pkg} generates [Nix]{.pkg} expressions that define reproducible environments 
  spanning [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system dependencies. Building on these environments, [rixpress]{.pkg} orchestrates polyglot pipelines where each 
  computational step runs in its own hermetically sealed environment, with automatic caching and dependency tracking. 
  This approach enables researchers to develop analyses interactively while maintaining bit-for-bit reproducibility, supports 
  collaboration across heterogeneous computational environments, and ensures that analyses remain executable years into the future.

@@

## Introduction: Reproducibility is also about software {#sec-intro}

@@

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, I interpret this notion more broadly as
the *computational environment*: the complete set of software required to
execute an analysis. Here too, a continuum exists. At the minimal end, authors
might only name the main software used—say, the [R]{.proglang} programming
language. More careful authors might also specify the version of [R]{.proglang},
or list the additional packages and their versions. Rarely, however, do authors
specify the operating system on which the analysis was performed, even though
differences in operating systems can lead to divergent results when using the
same code and software versions, as shown by @neupane2019. It is even less
common for authors to provide step-by-step installation instructions for the
required software stack.

@@

This paper focuses specifically on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. I present a comprehensive framework that addresses both
challenges using the [Nix]{.pkg} package manager, making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
their limitations to contextualize my contribution.

@@

However, [renv]{.pkg} has important limitations. It does not restore the version
of [R]{.proglang} itself—installing the correct R version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically,
[renv]{.pkg} does not handle system-level dependencies. If [sf]{.pkg} requires
[GDAL]{.pkg} version 3.0 but the system has version 2.4 installed, [renv]{.pkg} cannot
resolve this conflict. Users must manually install system libraries, and the
process differs across operating systems. Further details are discussed in the
[renv]{.pkg} documentation^[(https://rstudio.github.io/renv/articles/renv.html#caveats)].

@@

These tools represent significant progress in reproducibility, but they share a
common limitation: none addresses system-level dependencies or supports
polyglot environments. A project using [R]{.proglang},
[Python]{.proglang}, and system tools like [Quarto]{.pkg} or \LaTeX requires
coordinating multiple package managers ([renv]{.pkg} for R, virtual environments
for Python, manual installation for system tools), each with its own
configuration and potential for conflicts.

@@

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. Containers are typically run
non-interactively, allowing the complete computational environment to be
instantiated and executed with a single command.

@@

1. Poor interactive support: While containers can be modified at runtime,
  changes are lost when the container stops unless explicitly saved. Running
  graphical applications like RStudio Desktop from containers requires complex X11
  configuration that works reliably only on Linux. Instead, web-based IDEs (like RStudio
  Server) are used, but come with the inherent limitations of web-based solutions.

@@

4. Single-language focus: While [Docker]{.pkg} can support polyglot
  environments, orchestrating them requires manual coordination. There is no
  standard way to specify that one step should run in a Python environment while
  another runs in R.

@@

However, achieving this required mastering multiple complex tools: [renv]{.pkg}
for package management, [Docker]{.pkg} for environment containerization, and
[Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
and failure modes. The complexity compounds for polyglot projects; for instance,
a [Python]{.proglang} project would require learning virtual environments, 
and coordinating [R]{.proglang} and [Python]{.proglang} in a single analysis requires additional tooling.

@@

Together, these packages provide what existing tools cannot: a single framework
for managing reproducible, polyglot computational environments and
executing complex analytical workflows with step-level isolation. While
[Nix]{.pkg} remains complex for advanced use cases, [rix]{.pkg} and
[rixpress]{.pkg} abstract this complexity for common research workflows, making
deep reproducibility accessible without requiring systems administration
expertise.

The remainder of this paper proceeds as follows. Section 2 introduces the
[Nix]{.pkg} package manager and explains why its functional approach enables
reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
definition for various use cases. Section 4 discusses the `rstats-on-nix` fork
of the official package repository and how it addresses practical limitations.
Section 5 introduces [rixpress]{.pkg} and demonstrates polyglot pipeline
orchestration through a complete example. Section 6 concludes with discussion of
appropriate use cases and future directions.

@@

[Nix]{.pkg} is a powerful, cross-platform package manager designed for installing 
and building software in a fully reproducible and reliable manner. Unlike traditional 
package managers, [Nix]{.pkg} emphasizes immutable infrastructure and a functional
approach, which significantly enhances the consistency of computational environments.
As of writing, the primary Nix package collection, `nixpkgs`, provides access to over 
120,000 packages, including nearly all of CRAN and Bioconductor. According to Repology,
a service that tracks package repositories across distributions, `nixpkgs` is both the 
largest and among the most up-to-date package repositories available (see Figure @fig-repology).

![Package repository size and freshness according to Repology. `nixpkgs` leads in both 
total package count and update recency, making it well-suited for reproducible research 
environments.](repology.png){#fig-repology}

@@

For a more in-depth technical discussion of [Nix]{.pkg}'s design principles, I refer readers to @dolstra2004nix.

By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of 
tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files 
and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
polyglot environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
This enables the creation of a truly complete, project-specific, and deeply reproducible environment
that can be used interactively for development or non-interactively for automated analysis. 
As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
the environment can be rebuilt reliably in the future.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

Writing [Nix]{.pkg} expressions can be challenging for users unfamiliar with the
[Nix]{.proglang} language. However, the ability to define a fully reproducible
development environment in a single text file and then rebuild it anywhere is
highly appealing. To lower the barrier to adoption of [Nix]{.pkg} for
reproducibility, I developed the [rix]{.pkg} package.

@@

As explained earlier, [Nix]{.pkg} uses expressions from the `nixpkgs` GitHub
repository to build software. However, when generating expressions with
[rix]{.pkg}, the fork `rstats-on-nix/nixpkgs` is used instead.

@@

I strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to users of 
[rix]{.pkg} earlier than upstream, as merging PRs in the official repository can take some
time.

@@

A drawback of forking `nixpkgs` is that backported packages are not included
upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
to build many packages from source, which can be time-consuming. To mitigate
this, I provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
complementing the public Nix cache. Instructions for using Cachix are in
[rix]{.pkg}’s documentation. Using the cache significantly speeds up
installations, as prebuilt packages are downloaded rather than compiled.

@@

To illustrate these capabilities with a realistic research task, I present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang} [@bezanson2017julia], uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualizes the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code to this example can be found in the  
[following repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
using functions inspired by [targets]{.pkg}. The underlying logic for each step
is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
`functions.R`). Further details are provided as well as step-by-step instructions 
in the Appendix. The high-level orchestration script demonstrates how `{rixpress}` defines a
granular, multi-step machine learning workflow that crosses language boundaries:

@@

As a sidenote: [Python]{.proglang} users who wish to use [ryxpress]{.pkg} 
define pipelines using the same [R]{.proglang}-based DSL shown above. This design 
choice keeps the pipeline definition language consistent across both [R]{.proglang} 
and [Python]{.proglang} ecosystems. [ryxpress]{.pkg} calls [R]{.proglang} and 
[rixpress]{.pkg} under the hood to generate and build the `pipeline.nix` file.

@@

To execute the pipeline, one can either set `build = TRUE` in `rxp_populate()`
or call `rxp_make()` separately. [Nix]{.pkg} executes each step in order,
building dependencies as needed. Outputs are cached, so subsequent runs only
recompute steps with changed inputs or code. This provides the efficiency of
[targets]{.pkg} with polyglot support and bit-for-bit reproducibility.
Artifacts can be inspected interactively in [R]{.proglang} using
`rxp_read("artifact_name")` or `rxp_load("artifact_name")`.

@@

However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
[rix]{.pkg}, I aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
by providing a familiar interface and workflow. By building reproducible
development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
applications, or serving [plumber]{.pkg} APIs.

Furthermore, [rixpress]{.pkg} extends this reproducibility to entire analysis
pipelines. By leveraging Nix as a build automation engine, [rixpress]{.pkg}
allows polyglot workflows where each step runs in its own hermetically sealed
environment. This ensures deep reproducibility, efficient caching, and seamless
orchestration of polyglot analyses, making it easier to manage complex
projects with confidence that results can be reproduced exactly.

While [rix]{.pkg} and [rixpress]{.pkg} (and [ryxpress]{.pkg}) provide significant steps forward, it is 
important to acknowledge current limitations. 
[rixpress]{.pkg} (and [ryxpress]{.pkg}) in particular requires further development, especially to enhance
the ease of debugging complex pipelines. Its visualization capabilities are also still
fairly limited; for instance, they do not currently indicate outdated derivations when
a user modifies code that would necessitate a rebuild. These areas represent important 
avenues for future work to further mature the framework and improve the user experience.

@@

*   `functions/`: This directory contains the helper scripts with the core analytical code for each language.
    *   `functions.jl`: The [Julia]{.proglang} script for the economic simulation.
    *   `functions.py`: The [Python]{.proglang} script for the machine learning workflow.
    *   `functions.R`: The [R]{.proglang} script for data visualization.
*   `gen-env.R`: An [R]{.proglang} script that defines the reproducible computational environment.
*   `gen-pipeline.R`: The main [R]{.proglang} script that defines and orchestrates the entire polyglot pipeline.
